---
title: "Gov 50 Final Project"
author: Brock Locnikar
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts

-The data utilized for this project comes from the "CivilWar" data provided from class.

My research question is if greater than 1/3 export revenues from fuels is a causal factor in low polity scores within the latter half of the 20th century (Polity Score:  Index of democratization). We often hear about a supposed resource curse within the economies of dictatorial countries today. Often in the news, the apparent autocracies of the Middle East are blamed upon the prevalence of oil within their borders. However, I believe these claims must be verified by data. 

My hypothesis is that oil centralizes power structures and economies and this leads to the emergence of more autocratic governments with limited democratic institutions. The example often used for the petroleum producing curse is the Gulf region where despite high economic activity the region remained undemocratic throughout the 20th century. I believe that the effect oil has will be regional-agnostic and effect all regions of the world evenly. Meaning that all countries with greater than 1/3 export revenues from fuels will have lower polity scores than those that do not, especially within their own region. I hypothesize that the only reason that the Middle East is often synonymous with the "Oil Curse" is because this region has the most nations with oil. I do not believe that the undemocratic regimes (low polity scores) of MENA region nations is completely dependent on geography or culture of the Middle East. Rather, I hypothesize that these effects are solely from the economic dependence on oil.


I will be measuring the polity scores for Eastern Europe, Asia, Latin America, Western Democracies/Japan, Sub Saharan Africa, and the Middle East and North Africa for both when greater than 1/3 export revenues come from fuels and when less than 1/3 export revenues come from fuels. Polity scores measure the level of democratization of a state and will be a useful measure on the impact of oil production on governmental structures. 


If polity scores are higher for high oil producers in the same region than for low fuel producers, than my hypothesis could be incorrect that fuel production alone lowers democratization. If polity scores are lower for high fuel producers than for low fuel producers, than my hypothesis may be correct, but I will still have to run further analysis for confounding factors such as ongoing civil wars and GDP.


Thus, for the first data visualization I will be displaying how polity score (dependent variable) is affected by the production of oil. In order to maintain that the dependent variable is not impacted by other variables, the visualization also displays each major region of the earth. For instance, somewhere like the Middle East may be impacted by various outside factors that cause generally lower polity scores; however, measuring how the presence (or lack their of) of oil compared to other places in the same region is effective toward the understanding if the presence of oil alone is causal in lowering polity scores. Further, in order to have more in depth analysis, I will be measuring all locations from 1945 until 1991 when the Soviet Union collapsed. This will remeasure states each year through this time period, and this is helpful because oil reliance and polity can vary year to year. Thus, this combined fiugure can produce a much more accurate result.

Key Data and Visualization Points: Polity scores are measured on a scale from 10 to -10 with 10 being the most democratic and -10 being least democratic (most authoritarian). These scores are determined through expert evaluation of the government types. Furthermore, oil revenues are measured if oil makes up 1/3 or more of a nations GDP. Lastly, GDP per Pop is measured as GDP divided by the population. This bar plot displays the relationship between the variables polity2 score and oil (> 1/3 revenues from oil) - Japan was chosen to be grouped with Western European countries because of its advanced economy and European relationships throughout the 20th century.


```{r}
library(ggplot2)
library(dplyr)
library(gapminder)
library(tidyverse)


civilwar <- read_csv("~/Downloads/civilwar.csv")

plot1<- civilwar|>
  filter(year == 1945|1991) |>
  mutate(OilGeography = case_when(
      Oil == 0 & eeurop == 1  ~ "E. Europe Low Oil",
      Oil == 1 & eeurop == 1   ~ "E. Europe High Oil",
     Oil == 0 & lamerica == 1    ~ "Latin A. Low Oil",
     Oil == 1 & lamerica == 1    ~ "Latin A. High Oil",
     Oil == 0 & asia == 1     ~ "Asia Low Oil",
     Oil == 1 & asia == 1    ~ "Asia High Oil",
     Oil == 0 & nafrme == 1    ~ "MENA Low Oil",
     Oil == 1 & nafrme == 1   ~ "MENA High Oil",
     Oil == 0 & ssafrica == 1   ~ "SS Africa Low Oil",
     Oil == 1 & ssafrica == 1   ~ "SS AFrica High Oil",
     Oil == 0 & western == 1 ~ "West/Japan Low Oil",
     Oil == 1 & western == 1   ~ "West/Japan High Oil",
     TRUE ~ "Other"
    ))|>
  group_by(OilGeography)|>
  summarize(average_polity = mean(polity2, na.rm = TRUE))|>
  ggplot(mapping = aes(y = average_polity, x = OilGeography )) + geom_col(color = "blue") +
  theme(axis.text.x = element_text(angle = 90, size=8, vjust = 0.5, hjust=1)) + labs(y = "Average Polity Score", 
                                                                             x = "Region and Oil Dependency", title = "Average Polity Score Vs. Oil Revenue By Region", subtitle = "High Oil = Greater 1/3 Revenues From Fuel Exports")



plot1

civilwar



```
Discussion of Results: It seems, for some regions with higher than 1/3 of revenues oil production has a negative relation with polity scores. For instance, Asian, Eastern European, Middle East/North African, and Sub Saharan African countries with high oil revenues have lower polity scores than low oil producing states. However, for the West/Japan, and Latin American countries, high oil production is associated with higher polity scores. Thus, it seems oil appears to have a null effect on global polity scores.

However, since oil producing regions appear to have significant effects on regions that traditionally are associated with less economic activity, I believe it is important in the next visualization to determine if high oil revenues impact lower GDP per capita countries more than more economically affluent countries. GDP may be a confounding variable within the effects oil has on a nation. This data will also be performed on all nations and will include their repeating values from 1945-1991 in order to take into account regime and polity score shifts. 

Thus, 
```{r}
##pop ~ population, in 1000s


plot2<- civilwar|>
  filter(year == 1945|1991)|>
     group_by(country, gdpen, Oil)|>
  summarize(average_polity = mean(polity2, na.rm = TRUE))|>
  ggplot(mapping = aes(y = average_polity, x = gdpen, color = as.factor(Oil))) + geom_point() + scale_color_manual(values = c("0" ="blue", "1"= "red"))+
  theme(axis.text.x = element_text(angle = 90, size=8, vjust = 0.5, hjust=1)) + labs(y = "Average Polity Score", x = "Countries' GDP/POP Based on PWT 5.6, WDI 2001, and COW Energy Data  ", title = "Relationship Between GDP, Oil and Polity Scores") 
  

plot2

```
In contrast to what was believed earlier it seems that nations with both high GDP per capita and high oil production seem to have a correlation with lower average polity scores. Therefore, there may be confounding variables that have to do with regional location and how oil effects the polity scores. Thus, maybe nations with high production capacity from oil may be more effective in centralizing control of their government type away from the commons, and thus retaining lowered polity scores. Further, high GDP/pop may indicate that an extremely wealthy population may be okay with an authoritarian government because of high levels of disposable income. 




```{r}
civilwar45to91<- civilwar|>
  filter(year == 1945|1991)
  

Regression <- lm(polity2 ~ Oil,
               data = civilwar45to91)

Regression
```
Further, I chose to run a linear regression to determine how the presence of an oil reliant economy throughout all nations overall (from 1945 to 1991) has effects on polity scores. 
From this it is determined that the presence of Oil on average is correlated with on average -4.1181 lowered polity score. While, this regression does not factor in other variables such as variations between geography and national GDP, it is very interesting to see oil have a net drastic effect on the combinations of country values from 1945 to 1991. Further, this helps reinforce that oil alone has negative effects on democratization, often in conjunction with higher incomes.



Conclusion:

From the visualizations provided it is apparent that Oil has definitive effects on the polity scores of nations from 1945 to 1991. Nonetheless, these effects vary a lot by region and by GDP per capita of nations. There was some serious limitations with this project, as I did not have access to the exact oil revenue numbers for the years being investigated. Instead, I had to rely on the rough 1/3 revenue comparison. This disallowed extensive research into the contemporary era because many nations had shifted away from complete 1/3 revenues of oil, and thus, data was much more sparse. That is why I chose to document the years from 1945 to 1991. However, from this experiment, it is apparent that oil at least has some major effect on limiting democracy. However, for some reason Asian, Sub Saharan African, and Middle Eastern Countries are affected much more than their other global counterparts. Further, rather than being exasterbated by lower wealth, high GDP per capita countries experienced the lowest polity when paired with high oil consumption. This was very suprising to me. Therefore, in response to my initial hypothesis, it seems that oil limits democracy; however, this is very dependent on region and wealth levels of the nation. In the future, I would love to research how various other factors in the Middle East affect lowered democratic values coming from oil. Further, this research was limited by my inability to measure the same nations with and without effects from oil. This wouild be very useful in finding causality.


