---
title: "Gov 50 Final Project"
author: Brock Locnikar
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---


## Project thoughts

Data =  Civil War Data provided in class


My research question is if greater than 1/3 export revenues from fuels is a causal factor in low polity scores (Polity Score:  Index of democratization). 

My hypothesis is that oil centralizes power structures and economies and this leads to the emergence of more autocratic governments with limited democratic institutions. The example often used for the petroleum producing curse is the Gulf region where despite high economic activity the region remains undemocratic. I believe that the effect oil has will be regional-agnostic and effect all regions of the world evenly. Meaning that all countries with greater than 1/3 export revenues from fuels will have lower polity scores than those that do not, especially within their own region. 


I will be measuring the polity scores for Eastern Europe, Asia, Latin America, Western Democracies/Japan, Sub Saharan Africa, and the Middle East and North Africa for both when greater than 1/3 export revenues come from fuels and when less than 1/3 export revenues come from fuels. Polity scores measure the level of democratization of a state and will be a useful measure on the impact of oil production on governmental structures.git status



If polity scores are higher for high oil producers in the same region than low fuel producers, than my hypothesis is incorrect that fuel production alone lowers democratization. If polity scores are lower for high fuel producers than low fuel producers, then my hypothesis may be correct, but I will still have to run further anlysis for confounding factors such as ongoing civil wars and GDP.



polity2 score vs oil > 1/3 revenues from oil
```{r}
library(ggplot2)
library(dplyr)
library(gapminder)
library(tidyverse)


civilwar <- read_csv("~/Downloads/civilwar.csv")

plot1<- civilwar|>
  mutate(OilGeography = case_when(
      Oil == 0 & eeurop == 1  ~ "E. Europe Low Oil",
      Oil == 1 & eeurop == 1   ~ "E. Europe High Oil",
     Oil == 0 & lamerica == 1    ~ "Latin A. Low Oil",
     Oil == 1 & lamerica == 1    ~ "Latin A. High Oil",
     Oil == 0 & asia == 1     ~ "Asia Low Oil",
     Oil == 1 & asia == 1    ~ "Asia High Oil",
     Oil == 0 & nafrme == 1    ~ "MENA Low Oil",
     Oil == 1 & nafrme == 1   ~ "MENA High Oil",
     Oil == 0 & ssafrica == 1   ~ "SS Africa Low Oil",
     Oil == 1 & ssafrica == 1   ~ "SS AFrica High Oil",
     Oil == 0 & western == 1 ~ "West/Japan Low Oil",
     Oil == 1 & western == 1   ~ "West/Japan High Oil",
     TRUE ~ "Other"
    ))|>
  group_by(OilGeography)|>
  summarize(average_polity = mean(polity2, na.rm = TRUE))|>
  ggplot(mapping = aes(y = average_polity, x = OilGeography )) + geom_col(color = "blue") +
  theme(axis.text.x = element_text(angle = 90, size=8, vjust = 0.5, hjust=1)) + labs(y = "Average Polity Score", 
                                                                             x = "Region and Oil Dependency", title = "Average Polity Score Vs. Oil Revenue By Region", subtitle = "High Oil = Greater 1/3 Revenues From Fuel Exports")



plot1

civilwar



```
Discussion of Results: It seems, for some regions with higher than 1/3 of revenues oil production has a negative relation with polity scores. For instance, Asian, Eastern European, Middle East/North African, and Sub Saharan African countries with high oil revenues have lower polity scores than low oil producing states. However, for the West/Japan, Asian countries , and Latin American countries, high oil production is associated with higher polity scores. Thus, it seems oil appears to have a null effect on global polity scores.

However, since oil producing regions appear to have significant effects on regions that traditionally are associated with less economic activity, I believe it is important in the next visualization to determine if high oil revenues impact lower GDP per capita countries more than more economically affluent countries. GDP may be a confounding variable within the effects oil has on a nation. 

Thus, 
```{r}
##pop ~ population, in 1000s


plot2<- civilwar|>
     group_by(country, gdpen, Oil)|>
  summarize(average_polity = mean(polity2, na.rm = TRUE))|>
  ggplot(mapping = aes(y = average_polity, x = gdpen, color = as.factor(Oil))) + geom_point() + scale_color_manual(values = c("0" ="blue", "1"= "red"))+
  theme(axis.text.x = element_text(angle = 90, size=8, vjust = 0.5, hjust=1)) + labs(y = "Average Polity Score", x = "Countries' GDP/POP Based on PWT 5.6, WDI 2001, and COW Energy Data  ", title = "Relationship Between GDP, Oil and Polity Scores") 
  

plot2

```
In contrast to what was believed earlier it seems that nations with both high GDP per capita and high oil production seem to have a correlation with lower average polity scores. Therefore, there may be confounding variables that have to do with regional location and how oil effects the polity scores. In the next discussions I plan to delve deeper into how political unrest and Islamic adherence paired with highly oil dependent economies effect the lower polity scores in MENA, SS Africa, and Asia. Therefore, my initial hypothesis at this point seems inconclusive at best that oil productio alone negatively affects polity score. However, some regions to be affected more by high oil production.  

